Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-07-07 joint_inf.tex
Halp
oops, gibbs actually doesn't apply
notation and form of likelihood

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
%EM Is it indicated anywhere how the internal nodes are numbered? I would assume that 1 would be connected to leaf 1. If that's true please indicate here.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
%EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ is a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
By symmetry, exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
A similar property holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies, calculated using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}.
Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.

@@

    $(1+x)^2   (1+w)(1+y)^2$\\

@@

    $(1+x)(1-x)(1+w)(1+y)^2$\\

@@

    $(1+x)^2   (1+w)(1+y)(1-y)$\\

@@

    $\{\emptyset,\{1,2,3\}\}$&
    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\

@@

                    \{2\}\\

@@

                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1-x)^2   (1+w)(1+y)^2

@@

                $(1+x)^2   (1+w)(1+y)(1-y)$\\

@@

    $\emptyset$       &$\emptyset$&$(1+x)^2   (1+w)(1+y)^2$\\

@@

                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)

@@

                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2

@@

                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2

@@

                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1-x)^2   (1+w)(1+y)^2

@@

                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2

@@

@@

                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2

@@

p_0 = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})

@@

The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{2\}$ for the site split $\{1,3\}$.

@@

    &=        p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\

@@

    &\qquad + p_{123}\cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\

@@

    &\qquad + p_{123}\cdot\log((1+x')^2   (1+w')(1+y')(1-y')).

@@

(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y').

@@

    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\

@@

@@

    &\qquad + p_{123}\cdot\log((1+x')(1+y')(1-y')) \\
    &\qquad + p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\

@@

    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+x') \\

@@

    &\qquad + (1-p_{0}-p_{13})\cdot\log(1+y') \\

@@

    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\

@@

    &\qquad + (3-p_{0}-p_{13}-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\

@@

a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23},

@@

a_{3} = 3-p_{0}-p_{13}-p_{2}-p_{12}-p_{23}-p_{123},

@@

    &\le      p_{0}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\

@@

Applying Gibbs' inequality to the first two summands and maximizing over the remaining terms yields

@@

    &\le      p_{0}  \cdot\log(p_{0}) \\
    &\qquad + p_{13} \cdot\log(p_{13}) \\

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{0}  \cdot\log(p_{0}) \\
    &\qquad + p_{13} \cdot\log(p_{13}) \\
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
\end{align*}
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{0}  \cdot\log(p_{0}) \\
    &\qquad + p_{13} \cdot\log(p_{13}) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

@@

We construct a similar lower bound on the Felsenstein tree partial likelihood.
Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
Then, if $(1+x)(1-y) > (1-x)(1+y)$,
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{2}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{3}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{12} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{123}\cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
    &\qquad + p_{23} \cdot\log((1+x')^2   (1-w')(1-y')^2),
\end{align*}
otherwise
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{12} \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{123}\cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
    &\qquad + p_{13} \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{23} \cdot\log((1-x')^2    (1-w')(1+y')^2).
\end{align*}
Letting
$$
b = p_{1}+p_{2}+p_{3}+p_{123}
$$
yields
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(1+x') \\
    &\qquad + (2-b)  \cdot\log(1+y') \\
    &\qquad + b      \cdot\log(1-y') \\
    &\qquad + p_{0}\cdot\log(1+w') \\
    &\qquad + (1-p_{0})\cdot\log(1-w')
\end{align*}
for the first lower bound, but, upon maximizing, we obtain
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(2) \\
    &\qquad + (2-b)  \cdot\log(2-b) \\
    &\qquad + b      \cdot\log(b) \\
    &\qquad + p_{0}\cdot\log(2p_{0}) \\
    &\qquad + (1-p_{0})\cdot\log(2(1-p_{0}))
\end{align*}
maximized at $\{1,1-b,1-2p_{0}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
Our lower bound is
\begin{align*}
    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{0}\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{0}\cdot\log(2p_{0}) + (1-p_{0})\cdot\log(2(1-p_{0})).
\end{align*}

>>>>

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
%EM Is it indicated anywhere how the internal nodes are numbered? I would assume that 1 would be connected to leaf 1. If that's true please indicate here.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{2\}$ and internal node $\{2\}$ connected to leaves $\{3\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
%EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplit=\emptyset|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplit=\{1\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplit=\{3\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
$$
\left\{\begin{array}{l}
        (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2) \\
        (1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2) \\
        (1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2) \\
        (1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2)
\end{array}\right.
$$
for site split $\{1\}$ and
$$
\left\{\begin{array}{l}
        (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2) \\
        (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2) \\
        (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2) \\
        (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
\end{array}\right.
$$
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
By symmetry exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
A similar property holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies under these two topologies.
Table~\ref{tab:likelihoods} contains calculations of likelihood values for fixed site patterns and topologies that have been maximized over ancestral state patterns.
Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} contain the values of the likelihood for all possible ancestral states.

@@

    $(1+x')^2   (1+w')(1+y')^2$\\

@@

    $(1+x')(1-x')(1+w')(1+y')^2$\\

@@

    $(1+x')^2   (1+w')(1+y')(1-y')$\\

@@

    $\{\emptyset,\{1,2\}\}$&
    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\

@@

@@

                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1+w')(1-y')^2

@@

                $\{1,2\}$&
                $(1+x')^2   (1+w')(1+y')(1-y')$\\

@@

    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\

@@

                        (1+x')(1-x')(1+w')(1+y')^2\\
                        (1+x')^2   (1-w')(1+y')(1-y')

@@

                    (1+x')^2   (1+w')(1+y')(1-y')\\
                    (1+x')(1-x')(1-w')(1+y')^2

@@

                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1+x')^2   (1-w')(1+y')^2

@@

                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
                    (1+x')^2   (1+w')(1-y')^2

@@

                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1-x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1-w')(1-y')^2

@@

                    \{2\} \\

@@

                    (1+x')(1-x')(1-w')(1+y')^2 \\
                    (1+x')^2   (1+w')(1+y')(1-y')

@@

p_{\emptyset} = P(\siteSplitRV=\emptyset \mid \tau=\tau_1,t=\{x,y,y\})

@@

The full Farris tree log likelihood takes one of three values depending on branch lengths---assume for now the we are only interested in the ancestral state split $\{1\}$ for the site split $\{1,3\}$.

@@

    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\

@@

    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\

@@

    &\qquad + p_{123}        \cdot\log((1+x')^2   (1+w')(1+y')(1-y')).

@@

(1+x'^2-y'^2-x'^2y'^2) & = (1+x'^2)(1+y')(1-y') \le (1+x')(1+y')(1-y') \\
(1+x'^2+y'^2+x'^2y'^2) & = (1+x'^2)(1+y'^2) \le (1+x')(1+y') \\
4x'y' & = 2x' \cdot 2y' \le (1+x'^2)(1+y'^2) \le (1+x')(1+y')

@@

    &\le      p_{\emptyset}  \cdot\log(2(1+x')(1+y')) \\

@@

    &\qquad + p_{13}         \cdot\log((1+x')(1+y')) \\

@@

    &\qquad + p_{123}        \cdot\log((1+x')(1+y')(1-y')) \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\

@@

    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + \log(1+x') \\

@@

    &\qquad + \log(1+y') \\

@@

    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + (3-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\

@@

    &\qquad + (3-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\

@@

a_{1} = 3-p_{1}-p_{3}-p_{12}-p_{23},

@@

a_{3} = 3-p_{2}-p_{12}-p_{23}-p_{123},

@@

    &\le      p_{\emptyset}  \cdot\log(2) \\

@@

%EM I need help understanding this application of Gibbs. p_{0} + p_{13} is not 1 if I understand correctly, and so I don't see how Gibbs applies.
%You are totally right! That was an oversight on my part. I have redrawn the bounds on this term, and it will most likely affect the region of inconsistency, which is good, because I suspected something was making the inconsistency region larger than it should be. Though the bounds are not as tight as they could be now. Thanks!
Maximizing over the unknown terms yields

@@

    &\le      p_{\emptyset}  \cdot\log(2) \\

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
%EM Sorry to be slow, but could you provide a little more detail on why this follows? Is there some convexity of entropy argument going on here?
%Not sure which part is confusing, but:
    %1) a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
    %The below equation follows by rewriting 1+x and 1-x for the above x.
    %2) a_1' and a_2' are just defined by rewriting the likelihood. There are now two fewer log(1+x) terms and two more log(1-x) terms since we replaced a (1+x)^2 in the likelihood with a (1-x)^2. That's why we have a -2p_{13} and a +2p_{13} respectively.
    %Happy to clarify somewhere in the text but not sure what part would most need exposition.
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
\end{align*}
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

@@

We construct a similar lower bound on the Felsenstein tree partial likelihood.
Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
Then, if $(1+x)(1-y) > (1-x)(1+y)$,
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{3}          \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{12}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
    &\qquad + p_{23}         \cdot\log((1+x')^2   (1-w')(1-y')^2),
\end{align*}
otherwise
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      p_{\emptyset}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{2}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{12}         \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{123}        \cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
    &\qquad + p_{13}         \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{23}         \cdot\log((1-x')^2    (1-w')(1+y')^2).
\end{align*}
Letting
$$
b = p_{1}+p_{2}+p_{3}+p_{123}
$$
yields
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(1+x') \\
    &\qquad + (2-b)  \cdot\log(1+y') \\
    &\qquad + b      \cdot\log(1-y') \\
    &\qquad + p_{\emptyset}\cdot\log(1+w') \\
    &\qquad + (1-p_{\emptyset})\cdot\log(1-w')
\end{align*}
for the first lower bound, but, upon maximizing, we obtain
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &\ge      2\cdot\log(2) \\
    &\qquad + (2-b)  \cdot\log(2-b) \\
    &\qquad + b      \cdot\log(b) \\
    &\qquad + p_{\emptyset}\cdot\log(2p_{\emptyset}) \\
    &\qquad + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset}))
\end{align*}
maximized at $\{1,1-b,1-2p_{\emptyset}\}$, which is the same if $(1+x)(1-y) > (1-x)(1+y)$ or $(1+x)(1-y) < (1-x)(1+y)$.
Our lower bound is
\begin{align*}
    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{\emptyset}\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{\emptyset}\cdot\log(2p_{\emptyset}) + (1-p_{\emptyset})\cdot\log(2(1-p_{\emptyset})).
\end{align*}
